---
title: "Prj.1"
author: "Noah Hill"
date: "2022-10-30"
output: word_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Abstract

A Chinese automobile company Geely Auto aspires to enter the US market
by setting up their manufacturing unit there and producing cars locally
to give competition to their US and European counterparts. They have
contracted an automobile consulting company to understand the factors on
which the pricing of cars depends. Specifically, they want to understand
the factors affecting the pricing of cars in the American market, since
those may be very different from the Chinese market. From this
information we wanted to explore further and analyzed:

*What car features (ex: horsepower, citympg, highwaympg,
symbolling(insurance)) are most influential on the price of the car?* Is
there a correlation between type of car body and horse power?

# Introduction

We gathered this data from Kaggle.com. It has data of 205 different
individual cars from the US market. The variables are:

-   Symboling (Insurance Risk Rating) (-3=safe, +3=risky)
-   Car company
-   Fuel type (gas or diesel)
-   Aspiration (turbo or std)
-   Number of Doors
-   Car Body Type (convertible, hatchback, sedan, or hardtop)
-   Drive Wheel Type (4wd, rwd, or fwd)
-   Engine Location (front or rear)
-   Wheelbase in inches
-   Length of Car in inches
-   Width of Car in inches
-   Height of Car in inches
-   Weight of Car in pounds
-   Engine Type (dohc, ohc, ohcv, rotor, I, ohcf, dohcv,)
-   Number of Cylinders (\# ranges from 2-12)
-   Engine Size in cubic centimeters
-   Fuel System (1bbl, 2bbl, 4bbl, idi, mfi, mpfi, spdi, spfi)
-   Boreratio in mm
-   Stroke in mm
-   Compression Ratio in mm\^3
-   Horsepower
-   Peak RPM in thousands
-   City Miles per Gallon
-   Highway Miles per Gallon
-   Price of Car in dollars

Note that this data set is for learning purposes only, we cannot draw
any inference with real world scenarios. The response variable is the
price of the car. We are interested in determining which variables can
best predict the price of the car.

# Exploratory Analysis

## Check distributions of the predictor variables

```{r fig.height=7, fig.width = 8}
library(ggplot2)
library(tidyr)
ggplot(gather(CarPrice_Assignment [, c("price", "horsepower", "citympg", "highwaympg", "symboling")]), aes(value)) +
  geom_histogram(bins=10) + 
  facet_wrap(~key, scales = 'free_x')
```

The histogram of horsepower is right skewed. All other predictors have
reasonable distributions. We will consider a log transformation for
horsepower.

## Log Transformation for Horsepower variable

```{r}
sp = ggplot(gather(CarPrice_Assignment [, c("horsepower")]), aes(value)) +
  geom_histogram(bins=10) + 
  facet_wrap(~key, scales = 'free_x')
sp + scale_x_continuous(trans='log2') +
  scale_y_continuous(trans='log2')
```

After performing this transformation, it allowed us to remove skewness
and analyze it with a more normal distribution.

## Pairwise Correlations

```{r fig.height=9, fig.width = 9}
plot(CarPrice_Assignment [, c(10:14, 17, 19:26)])
fit2 = lm (price ~ log(horsepower, 10), data = CarPrice_Assignment)
summary(fit2)
```

Price is the most correlated with horsepower, city miles per gallon, and
highway miles per gallon. The correlation between price and horsepower
is 0.6196.

Our categorical predictor variables are symboling and carbody.

# Model 1

## First-order model with all predictors

```{r}
fit1 = lm (price ~ horsepower + citympg + highwaympg + carbody + symboling, data = CarPrice_Assignment)
x <- (data = CarPrice_Assignment [, c("price", "horsepower", "citympg", "highwaympg", "symboling")])
pair_data <- data.frame(x)
pairs(pair_data)
summary(fit1)
```

```{r}
library(corrplot)
cormat = cor (CarPrice_Assignment [, c(10:14, 17, 19:26)])
corrplot (cormat)
```

-   Using all predictor variables, the model explains 73.62% of the
    variation in price (adjusted ð‘…2).\
-   The residual standard error is \$4,186 (price).\
-   The most significant predictors are horsepower and
    carbodyhatchback.\
-   The predictor variables highwaympg and citympg show the highest
    correlation with each other. This is shown in the corr plot, the
    color is the darkest when comparing highwaympg and citympg.

## Residual Analysis - First-order model

```{r fig.width=7.5, fig.height=4}
par (mfrow=c(1,2))
plot(fit1)
```

-   There is curvature in the residuals.\
-   There is more constant residual variance in-between the fitted
    values when x = 15000 and x = 25000.
-   There is non-constant variance in the overall plot.
-   The Normal Q-Q plot shows more points above the line on the right
    side which means the residuals are not normally distributed. It has
    a longer tail on the right side.

## Box Cox Analysis

```{r}
library(MASS)
boxcox(fit1)
```

The Box-Cox analysis indicates that optimal transformation would have a
power below zero. Since zero is close to the 95% confidence interval for
ðœ†, we will use a log transformation.

# Stepwise Regression - First Order Model

## AIC

```{r}
step.all= step (fit1, direction="both")
summary (fit1)
```

The AIC regression didn't remove any predictor variables from the model.
The adjusted R\^2 value is 72.54% which suggests a high correlation for
our model without the removal of any predictors.

# Centering and Interaction Effects - Second Order Model

## a)

```{r}
#centering
mycenter = function (x) x - mean(x)

CarPrice_Assignment$horsepower.c = mycenter (CarPrice_Assignment$horsepower)

CarPrice_Assignment$citympg.c = mycenter (CarPrice_Assignment$citympg)

CarPrice_Assignment$highwaympg.c = mycenter (CarPrice_Assignment$highwaympg)

CarPrice_Assignment$highwarmpg.c = mycenter (CarPrice_Assignment$highwaympg)

#interaction effects
fit7 = lm (log(price) ~ (horsepower.c + citympg.c + highwaympg.c + carbody + symboling)^2, data = CarPrice_Assignment)
summary(fit7)
```

Through the interaction model and by centering, we concluded that the
most significant highly correlated varaibles are:
*citympg.c:highwaympg.c* citympg.c:symboling \*highwaympg.c:symboling

# Stepwise Regression - Second Order Model

## AIC

```{r}
step.all= step (fit7, direction="both")
summary (fit7)
```

Through the stepwise regression of AIC after centering the interaction
effects we found there to be the same three significant interactions
between the predictor variables. These include:\
*citympg:highwaympg* citympg:symboling \*highwaympg:symboling

The residual standard error is 0.214 with an adjusted R\^2 value of
0.8196.

# Final Model

## Residual Diagnostics

```{r}
plot(fit7)
```

Looking at the four residual plots there do not seem to be major
concerns. *Residuals vs fitted values plot: shows linearity and constant
variance conditions.* Normal QQ plot: values fall close to the line,
showing the residuals have a normal distribution. *Square root of
absolute standardized residuals vs fitted values: Shows there to be some
potential outliers (67,193,69) that we will analyze further. ASK ABOUT
THIS! something with +-3* Standardized residuals vs Leverage, with
Cook's distance: Cook's distance show no outliers to be greater than
0.5.

## Box Plot of Residuals

```{r}
boxplot(resid(fit7))
```

The box plot of the residuals show one major outlier with a high
standardized residual. WHAT ELSE DO I ADD?

## Plot of Response Variables vs. Fitted Values

```{r}
plot (log(price) ~ fit7$fitted.values, data=CarPrice_Assignment)
abline(0,1)
```

## Variance Inflation Factors

```{r}
library(car)
vif(fit7)
```

There are 5 VIF values that are less than 5, which is not great for the
number of variables and interactions it was analyzing. This shows a poor
collinearity assessment. We also found it strange that there were a
large numbers that could have been removed by AIC, so we notice this,
and may not conclude practicality from our results.

### High Leverage cutoff for fourth residual plot

```{r}
CarPrice_Assignment$leverage = hatvalues (fit7)

(lev.cut = 3 * fit7$rank / (fit7$rank + fit7$df.residual))

#plot (CarPrice_Assignment [, 2, 22, 24:26], pch=ifelse (CarPrice_Assignment$leverage > lev.cut, 2, 1))

```

The leverage cutoff is 0.4536585.

## Cook's Distance

```{r}
plot(fit7,which=5)
abline(v=lev.cut)
```

```{r}
summary(cooks.distance(fit7))
```

There are 5 rows (2.5% of the sample size) with leverage values above
the cutoff. We expect about 5%, so we have less outliers than a typical
data set. There are also no Cook's distance values above the cutoff of
0.5.

## Scatterplot Matrix

```{r}
plot(CarPrice_Assignment[,c("horsepower","citympg","highwaympg","carbody","symboling","price")],
     col=ifelse(hatvalues(fit7)>lev.cut,2,1))

table(CarPrice_Assignment$carbody)
```

The scatter plot matrix above shows the high leverage values as red
points that appear around the edges of the two-dimensional scatter
plots. We recognize that middle categories of symboling have high
leverage. We also see that car body has high leverage values around the
car body values of 1 and 2. We determined that R assigned labels to the
following: *1=convertible* 2=hardtop *3=hatchback* 4=sedan \*5=wagon
Based on this knowledge, we see that convertibles and hardtop cars have
the highest leverage values because there are fewer of these types of
cars in our data set.

## Interaction plots of the three most significant values

### citympg:highwaympg

```{r}
library(ggplot2)
library(dplyr)
library(ggpubr)
categorize = function (x) {
  quartiles = summary (x) [c(2, 3, 5)]
  result = rep ("Q1", length (x))
  result [(quartiles[1] < x) & (x <= quartiles [2])] = "Q2"
  result [(quartiles[2] < x) & (x <= quartiles [3])] = "Q3"
  result [quartiles[3] < x] = "Q4"
  return (result)
}
ggplot(data = CarPrice_Assignment, aes(x = citympg, y = log(price), 
                                       col = categorize(highwaympg))) +
  geom_point() +
  geom_smooth(method=lm)
```

The first interaction plot between city mpg and highway mpg shows the
most interesting results. To understand this graph, we analyzed that R
categorized highway mpg into four categories: *Q1=lowest 25% of highway
mpg values* Q2=25% to median of highway mpg values *Q3=median to 75% of
highway mpg values* Q4=highest 25% of highway mpg values With this
information we can conclude that the relationship between log(price) and
city mpg is stronger(steeper) for cars with low highway mpg values.

### citympg:symboling

```{r}
library(ggplot2)
library(dplyr)
library(ggpubr)

ggplot(data = CarPrice_Assignment, aes(x = citympg, y = log(price), col = as.factor(symboling))) +
  geom_point() +
  geom_smooth(method=lm)
```

The interaction plot between city mpg and symboling has less
statistically significant meaning because the slopes for the categories
of symboling (-2 to 3) and city mpg are all very close to similar.

### highwaympg:symboling

```{r}
library(ggplot2)
library(dplyr)
library(ggpubr)

ggplot(data = CarPrice_Assignment, aes(x = highwaympg, y = log(price), col = as.factor(symboling))) +
  geom_point() +
  geom_smooth(method=lm)
```

The interaction plot between highway mpg and symboling has less
statistically significant meaning because the slopes for the categories
of symboling (-2 to 3) and highway mpg are all very close to similar.

# Conclusion

The final model has an adjusted R\^2 = 0.8196, which means that 81.96%
of the variation in log sales price is explained by the model. The
residual standard error is 0.214 log dollars. Shown by our residual
analysis, this final model is consistent with the conditions for doing
linear regression.

Our most meaningful interaction effect was the relationship between city
mpg and highway mpg. The relationship between log(price) and city mpg is
stronger(steeper) for cars with low highway mpg values. DOES THIS MAKE
SENSE OR SHOULD BE IT HIGH HIGHWAY MPG VALUES?

MAKE EXAMPLE PREDICTIONS

ANSWER: Does your analysis raise any questions that can't be answered
from the current data set? If so, what are they?
